Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-10 LICS21/antitone.tex
Sec. 1: Added outline, "additional materials in appendices". Tidying things up, saving space.
Abstract: minor
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination. We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.

@@

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
Therefore
\begin{calculation}
\expect{Y_n \mid T > n}
   \step[=]{rearranging terms}
% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
% + 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
% - 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
\dfrac{
\begin{array}{l}
\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+ 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
\\
{} - 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
\end{array}
}{\mathbb P[T > n]}
  \step[=]{definition of conditional expectation}
\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{}
\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
  \step[\leq]{}
\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

@@

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{}
\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\begin{align*}
\expect{Y_{B_x}} 
&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
&\leq -1 < 0 
\end{align*}
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed PPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. 
Thanks to \Cref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$. 
Thus, by \Cref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}

\begin{restatable}{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

@@

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\[
f: {\Theta_2} \, n 
\mapsto 
g_2(n), \qquad 
0 \mapsto 0
\]
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to \Cref{thm:confluent ranking}, as can the others in this section with similar issues.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
  &=  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) \\
  &>  \frac 1 {n(n+2) + 1} \\
  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\end{align*}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

@@

Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages. (See e.g.~Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.)
Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
To our knowledge, the problem of proving AST of probabilistic programs with continuous distribution is new.

@@

\begin{example}[Escaping spline]
\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
\[
\underbrace{\big
(\tY \, f \, x . \,
0 \oplus_{\frac{1}{x+1}} f(x + 1) \big)}_{\Xi} 
\, 10
\]
\end{example}

@@

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\ F\ \underline x\ \underline y$ is antitone rankable.
\end{example}

>>>>

The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination. 
Thus we seek a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.

@@

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
Therefore
\begin{calculation}
\expect{Y_n \mid T > n}
   \step[=]{rearranging terms}
% \dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
% + 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
% - 
% \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
\dfrac{
\begin{array}{l}
\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+ 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
\\
{} - 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}
\end{array}
}{\mathbb P[T > n]}
  \step[=]{definition of conditional expectation}
\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{}
\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
%  \step[\leq]{}
%\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\mathbin{=} \dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

@@

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
%  \step[=]{}
\mathbin{=}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{}
\expect{\epsilon(Y_n) \cdot {\bf 1}_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
%\step[=]{}
\mathbin{=}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\begin{align*}
\expect{Y_{B_x}} 
&= \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})\\
&\leq -1 < 0 
\end{align*}
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n] \to 0$ as $n \to \infty$, therefore $T < \infty$ a.s.
\end{proof}

@@

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed PPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the probability space $(\entrosp, \Sigma_\entrosp, \mu_\entrosp)$, and define the same r.v.~$T_n, X_n, Y_n, T_M^{\tY}$. 
Thanks to \Cref{thm:rankable and strict rankable}, $(Y_n)_{n \geq 0}$ is a supermartingale; and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale w.r.t.~stopping time $T_M^{\tY}$. 
Thus, by \Cref{thm:a-c strict}, $T_M^{\tY} < \infty$ a.s.
\end{proof}

@@

Define an \emph{antitone sparse ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$:
\begin{inparaenum}[(i)]
\item $f(N) \geq 0$ for all $N \in \dom(f)$,
\item $M \in \dom(f)$,
\item for any $N \in \dom(f)$, evaluation of $N$ will eventually reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\iffalse
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\begin{restatable}{theorem}{AntitonePartial}  \label{thm:antitone partial implies rankable}
  Every antitone sparse ranking function is a restriction of an antitone ranking function.
\end{restatable}

@@

\begin{example}[Unbiased random walk]
\label{ex:unbiased random walk}
For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \ln(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone sparse ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
by
\(
f_2: {\Theta_2} \, n \mapsto 
g_2(n)
\)
and
$0 \mapsto 0$
for $n \in \mathbb N$.
% \begin{align*}
% {\Theta_2} \, n 
% &\mapsto 
% g_2(n)
% \\
% 0 &\mapsto 0.
% \end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$,\footnote{Technically this is not quite correct, because of the distinction between $\underline n + 1$ and $\underline{n + 1}$, but this ranking function can still be made to work by appealing to \Cref{thm:confluent ranking}, as can the others in this section with similar issues.} each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}%\textstyle
& g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 \\
  &=  \ln \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) 
  =  \frac 1 2 \ln\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \ln\left(1 + \frac 1 {n(n+2)}\right) 
  >  \frac 1 {n(n+2) + 1} \\
  &>  \frac 1 2 \frac 1 {(n+1)^2} + \frac 1 2 \frac 1 {(n+3)^2}
\end{align*}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone sparse ranking function, and $M_2$ is AST.
\end{example}

@@

Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages. (See e.g.~Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.)
Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
To our knowledge, the problem of proving AST of probabilistic functional programs with continuous distribution is new.

@@

\begin{example}[Escaping spline]
\label{ex:escaping spline}\cite[\S 5.4]{DBLP:journals/pacmpl/McIverMKK18}
\[
%\underbrace{
\big(\tY \, f \, x . \,
0 \oplus_{\frac{1}{x+1}} f(x + 1) 
\big)
%}_{\Xi} 
\, 10
\]
\end{example}

@@

\begin{example}[Higher-order recursion] \label{ex:higher-order recursion}
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \cite{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\qquad\tif{n \leq 0}{s}{
f \, n \, (\varphi \, f \, s \, (n - 1))
\oplus_p
f \, n \, (\varphi \, f \, s \, (n + 1))
}
\end{array}
\]
For any $F : \textsf{R} \to \textsf{R} \to \textsf{R}$ such that $F\ \underline n\ \underline m$ terminates with no $\tY$ reductions for all $m$ and $n$, $\Xi\ F\ \underline x\ \underline y$ is antitone rankable.
\end{example}

### 2021-01-10 LICS21/applications.tex
Abstract: minor
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems.\cite{kiselyov2016problems} Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. Selecting an appropriate addressing scheme can also be important in the context of variational inference.\cite[\S 6.2.2]{paige2016automatic}

>>>>

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. 
In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this labelling scheme turns out to have various problems \cite{kiselyov2016problems}. 
Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
Selecting an appropriate addressing scheme can also be important in the context of variational inference \cite[\S 6.2.2]{paige2016automatic}.

### 2021-01-10 LICS21/confluent.tex
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \emph{subterm of $M$ at a position $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\[
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\quad
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\]
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

@@

\begin{example} Consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$. If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to \dots \to N$ are independent from the samples for $M \to \dots \to N \to M \to \dots \to N$, and so on.
\end{example}

The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons. With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to \dots \to N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to \dots \to N_n \to O$.

@@

\begin{definition}
An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}

>>>>

A \emph{position} is a finite sequence of steps into a term, defined inductively as
\begin{align*}
\alpha ::= & \; {\cdot} \mid \lambda ; \alpha \mid @_1 ; \alpha \mid @_2 ; \alpha \mid \underline f_i ; \alpha \\
& {} \mid \tY ; \alpha \mid \textsf{if}_1 ; \alpha \mid \textsf{if}_2 ; \alpha \mid \textsf{if}_3 ; \alpha.
\end{align*}
The \emph{subterm of $M$ at $\alpha$}, denoted $M \mid \alpha$, is defined as
\begin{align*}
M \mid \cdot & = M \\
\lambda x. M \mid \lambda ; \alpha & = M \mid \alpha \\
M_1 M_2 \mid @_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2 \\
\underline f(M_1,\dots,M_n) \mid \underline f_i ; \alpha & = M_i \mid \alpha \quad \text{for }i \leq n \\
\tY M \mid \tY ; \alpha & = M \mid \alpha \\
\tif{M_1 < 0}{M_2}{M_3} \mid \textsf{if}_i ; \alpha & = M_i \mid \alpha \quad \text{for } i = 1,2,3
\end{align*}
so that every subterm is located at a unique position, but not every position corresponds to a subterm (e.g. $x \, y \mid \lambda$ is undefined). 
A position such that $M\mid \alpha$ does exist is said to \emph{occur} in $M$. 
%\changed[lo]{
\emph{Substitution} of $N$ at position $\alpha$ in $M$, written $M[N/\alpha]$, is defined similarly.
For example, let 
\[
M = \lambda x \, y. y \, (\tif{x < 0}{y \, (\underline f (x))}{\underline 3})
\quad
\alpha =\lambda ; \lambda ; @_2 ; \textsf{if}_2 ; @_2
\]
then 
\(
M[\tsample / \alpha] = \lambda x \, y. y \, (\tif{x < 0}{y \, \tsample}{\underline 3}).
\)
%}

@@

\begin{example} Consider the term $M = \tY (\lambda f x.\ \tif{\tsample - \underline{0.5} < 0}{f \, x}{x}) \, \underline 0$, which reduces after a few steps to $N = \tif{\tsample - \underline{0.5}  < 0}{M}{\underline 0}$. If the pre-selected sample for $(N,\textsf{if}_1;\underline{-}_1)$ is less than 0.5, $N$ reduces back to $M$, then $N$ again, then the same sample is used the next time, therefore it's an infinite loop, whereas if samples are labelled by reduction sequences, the samples for $M \to^\ast N$ are independent from the samples for $M \to^\ast N \to M \to^\ast N$, and so on.
\end{example}

The reduction sequences of skeletons will often be discussed as though they were just skeletons, identifying them with their final skeletons. With this abuse of notation, a reduction sequence $N$ (actually $N_1 \to^\ast N_n = N$) may be said to reduce to a reduction sequence $O$, where the reduction sequence implicitly associated with the final skeleton $O$ is $N_1 \to^\ast N_n \to O$.

@@

\begin{definition}
An \emph{antitone sparse ranking function on $M$ with respect to a reduction strategy $r$} is a partial function $f : Rch_r(M) \rightharpoonup \mathbb R$ such that there exists some antitone function $\epsilon : \nnReal \to \pReal$ such that
\begin{inparaenum}[(i)]
    \item $f(N) \geq 0$ for all $N \in \dom(f)$,
    \item $M \in \dom(f)$,
    \item for any $N \in \dom(f)$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}
\iffalse
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ at the positions specified by $r$ will eventually reach some $O$ such that either $r(O)$ isn't defined or $f(O)$ is, and $f(N) \geq \mathbb E[f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\end{definition}

### 2021-01-10 LICS21/ranking.tex
Reformatted def. of sparse ranking function (and two other similar def.), making it more compact.

%\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. 
\begin{definition}
Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually 
    %\akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability. Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.}
     reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\end{definition}
A sparse ranking function that is total is just a ranking function. Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

A similar procedure would work for other forms of syntactic sugar. If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined with respect to a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination. There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.

>>>>

%\paragraph{}
In both of these cases, there are only some values of the ranking function that are semantically important. 
\begin{definition}
Define a \emph{sparse ranking function} on a closed term $M$ to be a partial function $f : Rch(M) \rightharpoonup \mathbb R$ such that
\begin{inparaenum}[(i)]
    \item $f(N) \geq 0$ for all $N \in \dom(f)$,
    \item $M \in \dom(f)$,
    \item for any $N \in \dom(f)$, evaluation of $N$ will eventually 
    %\akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability. Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.}
     reach some $O$ which is either a value or in $\dom(f)$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of $\dom(f)$).
\end{inparaenum}

\iffalse
\begin{itemize}
    \item $f(N) \geq 0$ for all $N$ where $f$ is defined.
    \item $f$ is defined at $M$.
    \item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually 
    %\akr{It would be nicer for this to be only almost certain, to make this more neatly in-between rankability and AST, but that doesn't actually work because of terms reachable with 0 probability. Of course, ASTness doesn't imply rankability, so any definition of sparse ranking function that includes all Y-PAST terms wouldn't always extend to a total ranking function.}
     reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E[f(O) + \text{ the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}
\fi
\end{definition}
A sparse ranking function that is total is just a ranking function. Providing a sparse ranking function is essentially part way between providing a ranking function and directly proving almost sure termination.

@@

A similar procedure would work for other forms of syntactic sugar. If a closed term $N$ eventually reduces to one of a set of other terms $\{N_i \mid i \in I\}$ with certain probabilities, a sparse ranking function defined w.r.t.~a reduction sequence that skips straight from $N$ to $N_i$ is also a valid sparse ranking function for the original reduction function, and therefore its existence implies almost sure termination. There is a caveat, however, that $\tY$-reduction steps skipped over in the shortcut still need to be counted for the expected number of $\tY$-reduction steps.
